New approaches for boosting to uniformity

### 2014-10-13 assets/paperdraft/parts/introduction.tex
small textual edits

Methods of machine learning play an important role in modern particles physics. 
Multivariate classifiers, {\em e.g.}, boosted decision trees (BDTs) and artificial neural networks (ANNs), are now commonly used in analysis selection criteria. BDTs are now even used in software triggers~\cite{ref:lhcbhlt,ref:bbdt}. 
To enhance the performance, a series of classifiers is typically trained; this is a technique known as boosting.
Boosting  involves training many simple classifiers and then building a single composite classifier from their responses.
The classifiers are trained in series with the inputs of each member being augmented based on the performance of its predecessors.  This augmentation is designed such that each new classifier targets more those events which were poorly classified by previous members of the series.  The classifier obtained by combining all members of the series is typically much more powerful than any of the individual members.

In particle physics, the most common usage of BDTs is in classifying candidates as signal or background.  The BDT is determined by optimizing some figure of merit (FOM), {\em e.g.}, the signal purity or approximate signal significance.  This approach is optimal for a counting experiment; however, in many cases the BDT-based selection obtained in this way is not optimal.  
For example, in a Dalitz-plot (or any angular or amplitude analysis) analysis, obtaining a selection efficiency on signal candidates that is uniform across the Dalitz-plot is more important than any integrated FOM.  Similarly, when measuring a mean particle lifetime, obtaining an efficiency that is uniform in lifetime is what is desired.  In both cases, obtaining a uniform selection efficiency greatly reduces the systematic uncertainties involved in the measurement.  
When searching for a new particle, an analyst may want a uniform efficiency in mass for selecting background candidates so that the BDT-based selection does not generate a fake signal peak.  Furthermore, the analyst may also desire a uniform selection efficiency of signal candidates in mass (or other variates) since the new particle mass is not known.  In such cases, the BDT is often trained on simulated data generated with several values of mass (lifetime, {\em etc.}).  A uniform selection efficiency in mass ensures that the BDT is sensitive to the full range of masses involved in the search.

>>>>

Methods of machine learning play an important role in modern particles physics. 
Multivariate classifiers, {\em e.g.}, boosted decision trees (BDTs) and artificial neural networks (ANNs), are now commonly used in analysis selection criteria. BDTs are now even used in software triggers~\cite{ref:lhcbhlt,ref:bbdt}. 
To enhance the performance, a series of classifiers is typically trained; this is a technique known as boosting.
Boosting  involves training many simple classifiers and then building a single composite classifier from their responses.
The classifiers are trained in series with the inputs of each member being augmented based on the performance of its predecessors.  This augmentation is designed such that each new classifier targets those events which were poorly classified by previous members of the series.  The classifier obtained by combining all members of the series is typically much more powerful than any of the individual members.

In particle physics, the most common usage of BDTs is in classifying candidates as signal or background.  The BDT is determined by optimizing some figure of merit (FOM), {\em e.g.}, the signal purity or approximate signal significance.  This approach is optimal for a counting experiment; however, in many cases the BDT-based selection obtained in this way is not optimal.  
For example, in a Dalitz-plot (or any angular or amplitude analysis) analysis, obtaining a selection efficiency for signal candidates that is uniform across the Dalitz-plot is more important than any integrated FOM.  Similarly, when measuring a mean particle lifetime, obtaining an efficiency that is uniform in lifetime is what is desired.  In both cases, obtaining a uniform selection efficiency greatly reduces the systematic uncertainties involved in the measurement.  
When searching for a new particle, an analyst may want a uniform efficiency in mass for selecting background candidates so that the BDT-based selection does not generate a fake signal peak.  Furthermore, the analyst may also desire a uniform selection efficiency of signal candidates in mass (or other variates) since the mass of the new particle is not known.  In such cases, the BDT is often trained on simulated data generated with several values of mass (lifetime, {\em etc.}).  A uniform selection efficiency in mass ensures that the BDT is sensitive to the full range of masses involved in the search.

### 2014-10-13 assets/paperdraft/parts/metrics.tex
small textual edits

Figures~\ref{fig:peakdists} and~\ref{fig:pitdists} show the distribution of the predictions
and classifier efficiency as a function of mass for the peak and pit distributions, respectively. 
\begin{figure}[h]
\centering
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PeakDistributions.png}
			\caption{Mass vs prediction.}
		\end{subfigure}
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PeakEffs.png}
			\caption{Mass vs efficiency. The different coloured lines correspond to
				different global classifier efficiencies, as explained in the legend on the leftmost subplot.}
		\end{subfigure}
		\caption{Peak response distribution and efficiencies as a function of mass. \label{fig:peakdists}}
\end{figure}

\begin{figure}[h]
\centering
		% maybe we can just ignore this?
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PitDistributions.png}
			\caption{Mass vs prediction.}
		\end{subfigure}
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PitEffs.png}
			\caption{Mass vs efficiency. The different coloured lines correspond to
				different global classifier efficiencies, as explained in the legend on the leftmost subplot.}
		\end{subfigure}
		\caption{Pit response distribution and efficiencies as a function of mass. \label{fig:pitdists}}
\end{figure}

\begin{figure}[h]
\centering
	\includegraphics[width=0.9\textwidth]{graphs/TheilVsMSE.png}
	\caption{Comparison of SDE and Theil uniformity measures for different values of the constant $\alpha$ for the peak and pit distributions.
The Theil can clearly distinguish between these two while the SDE measure cannot.}
\end{figure}

@@

\subsubsection*{$D \to hhh$}
When our classifiers where applied to data set $D \to hhh$ we got uniformity measures for the SDE and Theil
metrics shown in Fig.~\ref{fig:d2hhhsdetheil}.
Both metrics show similar results, so there is no significant difference between them, nor the CvM metric, on this dataset.

\begin{figure}[h]
\centering
		% TODO use metrics on signal for signal, use metrics on bck for bck.
		\begin{subfigure}[b]{0.48\textwidth}
			\includegraphics[width=\textwidth]{graphs/D23hSignalEffs.png}
			\caption{Metrics for $D \to hhh$ signal.}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\includegraphics[width=\textwidth]{graphs/D23hBgEffs.png}
			\caption{Metrics for $D \to hhh$ background.}
		\end{subfigure}
		\caption{Uniformity metrics for the $D \to hhh$ data set. The (top) signal and (bottom) background uniformities are plotted as a function of the training stage of a given classifier, listed in the legends on the leftmost plots. \label{fig:d2hhhsdetheil}}
\end{figure}

>>>>

Figures~\ref{fig:peakdists} and~\ref{fig:pitdists} show the distribution of the predictions
and classifier efficiency as a function of mass for the peak and pit distributions, respectively. 
\begin{figure}[h]
\centering
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PeakDistributions.png}
			\caption{Mass vs prediction.}
		\end{subfigure}
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PeakEffs.pdf}
			\caption{Mass vs efficiency. The different coloured lines correspond to
				different global classifier efficiencies, as explained in the legend on the leftmost subplot.}
		\end{subfigure}
		\caption{Peak response distribution and efficiencies as a function of mass. \label{fig:peakdists}}
\end{figure}

\begin{figure}[h]
\centering
		% maybe we can just ignore this?
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PitDistributions.png}
			\caption{Mass vs prediction.}
		\end{subfigure}
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/PitEffs.pdf}
			\caption{Mass vs efficiency. The different coloured lines correspond to
				different global classifier efficiencies, as explained in the legend on the leftmost subplot.}
		\end{subfigure}
		\caption{Pit response distribution and efficiencies as a function of mass. \label{fig:pitdists}}
\end{figure}

\begin{figure}[h]
\centering
	\includegraphics[width=0.9\textwidth]{graphs/TheilVsMSE.pdf}
	\caption{Comparison of SDE and Theil uniformity measures for different values of the constant $\alpha$ for the peak and pit distributions.
The Theil can clearly distinguish between these two while the SDE measure cannot.}
\end{figure}

@@

\subsubsection*{$D \to hhh$}
Finally, we compare the SDE and Theil uniformity measures for the classifiers applied to the $D \to hhh$ data set, as shown in Fig.~\ref{fig:d2hhhsdetheil}.
Both measures show similar results, so there is no significant difference between them, nor the CvM metric, on this dataset.

\begin{figure}[h]
\centering
		% TODO use metrics on signal for signal, use metrics on bck for bck.
		\begin{subfigure}[b]{0.48\textwidth}
			\includegraphics[width=\textwidth]{graphs/D23hSignalEffs.pdf}
			\caption{Metrics for $D \to hhh$ signal.}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\includegraphics[width=\textwidth]{graphs/D23hBgEffs.pdf}
			\caption{Metrics for $D \to hhh$ background.}
		\end{subfigure}
		\caption{Uniformity metrics for the $D \to hhh$ data set. The (top) signal and (bottom) background uniformities are plotted as a function of the training stage of a given classifier, listed in the legends on the leftmost plots. Within each figure the leftmost plot is the SDE and the rightmost plot is the Theil uniformity metric. \label{fig:d2hhhsdetheil}}
\end{figure}

### 2014-10-13 assets/paperdraft/ugb.tex
small textual edits

The gradient boosting~\cite{ref:F1999} (GB) algorithm category requires the analyst to choose a differentiable loss function with the goal of building a classifier that minimizes the loss.  
A popular choice of loss function is the so-called AdaLoss function 
\begin{equation}
L_{\rm ada} = \sum\limits_{i=1}^{n} {\rm exp}\left[-\gamma_i s_i\right]. 
\end{equation}
The {\em scores} $s$ are obtained for each event as sum of predictions of all elements in the series. 
At each stage in the gradient boosting process, a {\em regressor} (a decision tree in our case) is trained whose purpose is to decrease the loss.  This is accomplished using the gradient decent method and the {\em pseudo-residuals}
\begin{equation}
  -\frac{\partial L_{\rm ada}}{\partial s_i} = \gamma_i  {\rm exp}\left[-\gamma_i s_i\right],
\end{equation}
which are positive(negative) for signal(background) events and have larger moduli for poorly classified events.

@@

Another approach is to include in the definition of the loss function some uniformity metric.  Consider first the case where the data have been binned in \y.  If the distribution of classifier responses in each bin, $f_b(s)$, is the same as the global response distribution, $f(s)$, then any cut made on the response will produce a uniform selection efficiency in \y.  Therefore, performing a one-dimensional goodness-of-fit test of the hypothesis $f_b \equiv f$ in each bin provides an assessment of the selection uniformity.  
For example, one could perform the Kolmogorov-Smirnov test in each bin and define a loss function as follows:
\begin{equation}
  L_{\rm flat(KS)} = \sum\limits_{b} w_b {\rm max}|F_b(s)-F(s)|,
\end{equation}
where $F_{(b)}(s)$ denotes the cumulative distribution of $f_{(b)}(s)$ and $w_b = \sum \delta({\rm bin}_i - b)/n_{\rm signal}$, {\em i.e.} $w_b$ is the fraction of signal events in the bin\footnote{If weighted events are used, then the fractional sum of weights should be used for $w_b$.}.

The gradient of the Kolmogorov-Smirnov loss function is zero for events with responses greater than the value of $s$ at which ${\rm max}|F_b(s)-F(s)|$ occurs.  Thus, it is not suitable for gradient boosting due to its instability.  Instead, we use the following {\em flatness} loss function:
\begin{equation}
  L_{\rm flat} = \sum\limits_{b} w_b \int |F_b(s)-F(s)|^2 ds,
\end{equation}
whose pseudo-residuals are ($b$ is the bin containing the $k$th event)
\begin{equation}
 -\frac{\partial L_{\rm flat}}{\partial s_k} \approx - 2 w_b \left[F_b(s_k)-F(s_k)\right] .
\end{equation}
This so-called flatness loss penalizes non-uniformity but does not consider the quality of the classification.  Therefore, the full loss function used is 
\begin{equation}
\label{eq:L_ada_flat}
  L_{{\rm ada}+{\rm flat}} = L_{\rm flat} + \alpha L_{\rm ada},
\end{equation}  
where $\alpha$ is a real-valued parameter that is typically chosen to be small.  The first term in Eq.~\ref{eq:L_ada_flat} penalizes non-uniformity, while that second term penalizes poor classification.  
We refer to this algorithm as uGB with flatness loss (uGBFL). 
In principle, many different flatness loss functions can be defined and could be substituted for our choice here.  See Appendix~A for a detailed discussion on this topic.

>>>>

The gradient boosting~\cite{ref:F1999} (GB) algorithm category requires the analyst to choose a differentiable loss function with the goal of building a classifier that minimizes the loss.  
A popular choice of loss function is the so-called AdaLoss function 
\begin{equation}
L_{\rm ada} = \sum\limits_{i=1}^{n} {\rm exp}\left[-\gamma_i s_i\right]. 
\end{equation}
The {\em scores} $s$ are obtained for each event as the sum of predictions of all elements in the series. 
At each stage in the gradient boosting process, a {\em regressor} (a decision tree in our case) is trained whose purpose is to decrease the loss.  This is accomplished using the gradient decent method and the {\em pseudo-residuals}
\begin{equation}
  -\frac{\partial L_{\rm ada}}{\partial s_i} = \gamma_i  {\rm exp}\left[-\gamma_i s_i\right],
\end{equation}
which are positive(negative) for signal(background) events and have larger moduli for poorly classified events.

@@

Another approach is to include some uniformity metric in the definition of the loss function.  Consider first the case where the data have been binned in \y.  If the distribution of classifier responses in each bin, $f_b(s)$, is the same as the global response distribution, $f(s)$, then any cut made on the response will produce a uniform selection efficiency in \y.  Therefore, performing a one-dimensional goodness-of-fit test of the hypothesis $f_b \equiv f$ in each bin provides an assessment of the selection uniformity.  
For example, one could perform the Kolmogorov-Smirnov test in each bin and define a loss function as follows:
\begin{equation}
  L_{\rm flat(KS)} = \sum\limits_{b} w_b {\rm max}|F_b(s)-F(s)|,
\end{equation}
where $F_{(b)}(s)$ denotes the cumulative distribution of $f_{(b)}(s)$ and $w_b = \sum \delta({\rm bin}_i - b)/n_{\rm signal}$, {\em i.e.} $w_b$ is the fraction of signal events in the bin\footnote{If weighted events are used, then the fractional sum of weights should be used for $w_b$.}.

The gradient of the Kolmogorov-Smirnov loss function is zero for events with responses greater than the value of $s$ at which ${\rm max}|F_b(s)-F(s)|$ occurs.  Thus, it is not suitable for gradient boosting due to its instability.  Instead, we use the following {\em flatness} loss function:
\begin{equation}
  L_{\rm flat} = \sum\limits_{b} w_b \int |F_b(s)-F(s)|^2 ds,
\end{equation}
whose pseudo-residuals are ($b$ is the bin containing the $k$th event)
\begin{equation}
 -\frac{\partial L_{\rm flat}}{\partial s_k} \approx - 2 w_b \left[F_b(s_k)-F(s_k)\right] .
\end{equation}
This so-called flatness loss penalizes non-uniformity but does not consider the quality of the classification.  Therefore, the full loss function used is 
\begin{equation}
\label{eq:L_ada_flat}
  L_{{\rm ada}+{\rm flat}} = L_{\rm flat} + \alpha L_{\rm ada},
\end{equation}  
where $\alpha$ is a real-valued parameter that is typically chosen to be small.  The first term in Eq.~\ref{eq:L_ada_flat} penalizes non-uniformity, while the second term penalizes poor classification.  
We refer to this algorithm as uGB with flatness loss (uGBFL). 
In principle, many different flatness loss functions can be defined and could be substituted for our choice here.  See Appendix~A for a detailed discussion on this topic.
